\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-06-13 doc/incertae_sedis.tex
Update.

\newcommand{\mthcomment}[1]{{\color{red} #1}\xspace}

\newcommand{\bdrcomment}[1]{{\color{cyan} #1}\xspace}

@@

\paragraph{Equivalence of rooted split definitions with character encodings}
Representing a split in a tree by a character is the basis of one of the oldest
supertree methods: matrix representation with parsimony \citep[MRP, see][]{Baum1992,Ragan1992}.
In the rooted form of such an approach, character state 1 is used for tips that are
descendants of a internal node in an input tree and 0 is used for other tips and for the root.
The partial splits caused by \emph{incertae sedis} annotations naturally map to
an encoding of missing data (typically denoted with the `?' symbol in a matrix).
Thus the taxonomy shown in Figure \ref{fig:Handling-incertae-sedis}b could be represented as the matrix shown in 
Figure \ref{fig:Handling-incertae-sedis-mrp}.
In the naive treatment of ignoring the \emph{incertae sedis} annotation, the `?' characters
   in the row for taxa $C1$, $C2$, and $C3$ would be replaced with 0's.

\begin{figure}
\begin{center}
\begin{tabular}{|r|p{2em}|p{2em}|p{2em}|p{2em}|}
\hline
& \multicolumn{3}{|c|}{Character/clade label} \\
leaves & C & AB & A \\
\hline
$A1$       & 0 & 1 & 1 \\
$A2$       & 0 & 1 & 1 \\
$B$        & 0 & 1 & 0 \\
$C1$       & 1 & ? & ? \\
$C2$       & 1 & ? & ? \\
  $D$      & 0 & 0 & 0 \\
  root     & 0 & 0 & 0 \\
\hline
\end{tabular}
\end{center}
\caption{\label{fig:Handling-incertae-sedis-mrp} The matrix encoding 
of an \emph{incertae sedis} interpretation of the taxonomy shown in  Figure \ref{fig:Handling-incertae-sedis}b.
The root is coded as a special leaf.
}

@@

An \emph{incertae sedis} taxon $n$ differs from a normal taxon because it can be
placed as a descendant of some set of taxa without invalidating these taxa.
We will use $\mathcal{P}_n$ to denote this set of taxa which are {\emph{potential}} 
ancestors for \emph{incertae sedis} taxon $n$.
Note that this distinct from the taxa that are ancestral to $n$ in the taxonomy;
    those are taxa which must contain all of the descandants of $n$.
The effect of the \emph{incertae sedis} annotation of taxon $n$ is
    to simply remove $\mathcal{I}(n)$ from the exclude set for each $p\in\mathcal{P}_n$.
The include sets of these splits remain unchanged, but the exclude sets may be reduced
\emph{incertae sedis} taxa to intrude.
Such a split does
not mention the complete set of tip labels $\mathcal{L}$ and is therefore called a
`partial split'', as opposed to a ``full split'' that contains the entirely of $\mathcal{L}$.

@@

\newpage %TEMP

>>>>

%\newcommand{\mthcomment}[1]{{\color{red} #1}\xspace}
\newcommand{\mthcomment}[1]{}

%\newcommand{\bdrcomment}[1]{{\color{cyan} #1}\xspace}
\newcommand{\bdrcomment}[1]{}

@@

\paragraph{Equivalence of rooted split definitions with character encodings}
Representing a split in a tree by a character is the basis of one of the oldest
supertree methods: matrix representation with parsimony \citep[MRP, see][]{Baum1992,Ragan1992}.
In the rooted form of such an approach, character state 1 is used for tips that are
descendants of a internal node in an input tree and 0 is used for other tips and for the root.
Partial splits, such as those caused by \emph{incertae sedis} annotations, are naturally encoded by
encoding missing labels in a split as missing data in the corresponding character
(typically denoted with the `?' symbol in a matrix).

A rooted split is equivalent to the corresponding character in a matrix representation in the
sense that each can be treated as a condition that each tree may or may not satisfy.  A tree satisfies
a split if there a branch which divides include and exclude group when cut.  A tree satisfies the
corresponding character in a matrix representation if the character has a parsimony score of $1$ on the tree.  This equivalence is illustrated in Figure \ref{fig:Handling-incertae-sedis-mrp}.  Thus the taxonomy shown in Figure \ref{fig:Handling-incertae-sedis}b could be represented in terms of the splits
in Figure \ref{fig:Handling-incertae-sedis-mrp}a or the characters in Figure \ref{fig:Handling-incertae-sedis-mrp}b.  In the naive treatment of ignoring the \emph{incertae sedis} annotation, the `?' characters in the row for taxa $C1$ and $C2$ would be replaced with 0's.  Equivalently, the naive interpretation would add $C1$ and $C2$ to the exclude set for the $A$ and $AB$ splits.

   \begin{figure}
     \subfloat[Split representation]{
\noindent\begin{minipage}[t]{0.45\columnwidth}%
\begin{align*}
C & =C1\,C2\,\phantom{B\,}|\bullet\,A1\,A2\,B\,D\\
A & =A1\,A1\,\phantom{B\,}|\bullet\,B\,D\\
AB & =A1\,A2\,B\,|\bullet\,D
\end{align*}
%
\end{minipage}

}
     \subfloat[Matrix representation]{
         \begin{tabular}{|r|p{2em}|p{2em}|p{2em}|p{2em}|}
           \hline
           & \multicolumn{3}{|c|}{Character/clade label} \\
           leaves & C & AB & A \\
           \hline
           $A1$       & 0 & 1 & 1 \\
           $A2$       & 0 & 1 & 1 \\
           $B$        & 0 & 1 & 0 \\
           $C1$       & 1 & ? & ? \\
           $C2$       & 1 & ? & ? \\
           $D$      & 0 & 0 & 0 \\
           (root)     & 0 & 0 & 0 \\
           \hline
         \end{tabular}
}
       \caption{\label{fig:Handling-incertae-sedis-mrp} Rooted splits have an equivalent matrix encoding.
        (a) Rooted splits for an \emph{incertae sedis} interpretation of the taxonomy shown in  Figure \ref{fig:Handling-incertae-sedis}b. (b) The equivalent matrix representation. The root is coded as a special leaf. Split C is a full split, because it mentions all labels.  Splits A and AB are partial splits, and therefore have missing labels in the left panel, and '?' in the right panel.
       }

@@

An \emph{incertae sedis} taxon $n$ differs from a normal taxon because it can be
placed as a descendant of some set of taxa without invalidating these taxa.
We will use $\mathcal{P}_n$ to denote this set of taxa which are {\emph{potential}} 
ancestors for \emph{incertae sedis} taxon $n$.
Note that this is distinct from the taxa that are ancestral to $n$ in the taxonomy;
    those are taxa which must contain all of the descandants of $n$.
The effect of the \emph{incertae sedis} annotation of taxon $n$ is
    to simply remove $\mathcal{I}(n)$ from the exclude set for each $p\in\mathcal{P}_n$.
The include sets of these splits remain unchanged, but the exclude sets may be reduced
\emph{incertae sedis} taxa to intrude.
Such a split does
not mention the complete set of tip labels $\mathcal{L}$ and is therefore called a
`partial split'', as opposed to a ``full split'' that contains the entirely of $\mathcal{L}$.

@@

% \newpage %TEMP
